Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-05-09 analysis/paper/paper.qmd
qualitative studies and repro

Although the introduction of reproducibility reviews signifies a growth in computational archaeology and a desire to evaluate the research products beyond the journal article, a very substantial amount of archaeological research is qualitative, with few or no numerical data involved in making knowledge claims about the human past. For example, many archaeological questions can be answered by the simple presence or absence of artefacts or features, or qualitative comparisons of basic artefact characteristics such as shape, colour, surface treatments, and raw material. Chaîne opératoire analyses by ceramic and lithic specialists is an especially productive area of archaeological research that often relies on comparison of narratives of manufacturing processes. Do these studies count as science and should they be held to standards of reproducibility in the same way that computationally intensive research is?

While there is a strong case for qualitative research counting as science, because it is often a systematic, empirical, and rigorous process of inquiry, the question of

>>>>

Although the introduction of reproducibility reviews signifies a growth in computational archaeology and a desire to evaluate the research products beyond the journal article, a very substantial amount of archaeological research is qualitative, with few or no numerical data involved in making knowledge claims about the human past. For example, many archaeological questions can be answered by the simple presence or absence of artefacts or features, or qualitative comparisons of basic artefact characteristics such as shape, colour, surface treatments, and raw material. Chaîne opératoire analyses by ceramic and lithic specialists is an especially productive area of archaeological research that often relies on comparison of narratives of manufacturing processes. While these studies unquestionably count as science, because they are a systematic, empirical, and rigorous process of inquiry, should they be held to standards of reproducibility in the same way that computationally intensive research is?

A similar debate has been unfolding more broadly about the humanities where Peels and Bouter [-@peelsHumanitiesNeedReplication2018; -@peelsPossibilityDesirabilityReplication2018] have argued that humanities disciplines that use empirical methods should be assessed by how well reanalysis of the original or new data using original or new methods produces the original or equivalent results. Resisting this proposal, @derijckeResistCallsReplicability2018 argue that because humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, replication is irrelevant as a mark of quality. @peelsEpistemicValuesHumanities2018 disputes this uniqueness, claiming that the humanities has the same epistemic values as the sciences, however some values have more weight in the humanities while others have more weight in the sciences, and unlike in the sciences, humanities scholars often study these epistemic values themselves. A consensus seems to be emerging that
